Data organization in spreadsheets

### 2017-06-06 manuscript.Rmd
Further changes from AE comments

@@

_Use consistent codes for categorical variables._ For a
categorical variable like *sex*, use a single common value for
males (e.g. "`male`") and a single common value for females
(e.g. "`female`"). Don't sometimes write "`M`", sometimes "`male`",
and sometimes "`Male`". Pick one and stick to it.

@@

_Use a consistent format for all dates,_ preferably with the standard
format `YYYY-MM-DD`, for example `2015-08-01`. If sometimes you write `8/1/2015`
and sometimes `8-1-15`, you are asking for trouble.

@@

_Be careful about extra spaces within
cells._ A blank cell is different than a cell that contains a single space. And
"`male`" is different from "` male `" (that is, with spaces at the beginning and
end). These can be a headache later on.

@@

Be careful about extraneous spaces (say, at the beginning or end of a
variable name). "`glucose`" is different from "`glucose `" (with an
extra space at the end). Extraneous spaces can cause headaches later.

@@

But note that Microsoft Excel does bizarre things with dates
(see <https://storify.com/kara_woo/excel-date-system-fiasco>). It stores them
internally as an number, counting the days since 1900-01-01. Wait, that is only
for Windows; on Macs, it counts the days since 1904-01-01. So, be careful to
check that the dates haven't been mangled when your data come out of Excel.

@@

Figure 1 displays a portion of a spreadsheet that we got from a
collaborator. We don't quite remember what those e's were for, but in
any case having
different date formats within a column can cause headaches later.

@@

You could fill in some of those cells, to make it more clear, but even
better would be to make a "tidy" version of the data [@wickham2014],
with each row being one replicate and with the response values all in one column, as in
Figure 3. We will discuss this in more detail, below.

@@

Even better would be to make it a "tidy" data set
with each row being a subject on a
particular day, as shown in Figure 8.

@@

Another possible use of highlighting would be to indicate males and
females by highlighting the corresponding rows in different
colors. But rather than use highlighting to indicate sex, it is better
to include a `sex` column, with values `Male` or `Female`.

@@

Lance Waller and Lincoln Mullen generously provided comments to improve the manuscript.

>>>>

Murrell [-@murrell2013] contrasts data that is formatted for humans to
view by eye with data that is formatted for a computer. He provides an
extended example of computer code to extract data from a set of files
with complex arrangements. It is important that data analysts be able
to work with such complex data files. But if the initial arrangement
of the data files is planned with the computer in mind, this would
simplify the later analysis process.

@@

_Use consistent codes for categorical variables._ For a
categorical variable like the sex of a mouse in a genetics study, use a single common value for
males (e.g. "`male`") and a single common value for females
(e.g. "`female`"). Don't sometimes write "`M`", sometimes "`male`",
and sometimes "`Male`". Pick one and stick to it.

@@

_Use a consistent format for all dates,_ preferably with the standard
format `YYYY-MM-DD`, for example `2015-08-01`. If sometimes you write `8/1/2015`
and sometimes `8-1-15`, it will be more difficult to use the dates in
analyses or data visualizations.

@@

_Be careful about extra spaces within
cells._ A blank cell is different than a cell that contains a single space. And
"`male`" is different from "` male `" (that is, with spaces at the beginning and
end).

@@

Be careful about extraneous spaces (say, at the beginning or end of a
variable name). "`glucose`" is different from "`glucose `" (with an
extra space at the end).

@@

But note that Microsoft Excel does bizarre things with dates
(see <https://storify.com/kara_woo/excel-date-system-fiasco>). It stores them
internally as an number, counting the days since 1900-01-01. Wait, that is only
for Windows; on Macs, it counts the days since 1904-01-01. So, you may
need to manually check that the dates haven't been mangled when your data
come out of Excel.

@@

Figure 1 displays a portion of a spreadsheet that we got from a
collaborator. We don't quite remember what those e's were for, but in
any case having
different date formats within a column makes it more difficult to use
the dates in later analyses or data visualizations.

@@

You could fill in some of those cells, to make it more clear.
Alternatively, make a "tidy" version of the data [@wickham2014],
with each row being one replicate and with the response values all in one column, as in
Figure 3. We will discuss this further, below.

@@

Alternatively, make it a "tidy" data set
with each row being a subject on a
particular day, as shown in Figure 8.

@@

Another possible use of highlighting would be to indicate males and
females in a mouse study by highlighting the corresponding rows in different
colors. But rather than use highlighting to indicate sex, it is better
to include a `sex` column, with values `Male` or `Female`.

@@

Lance Waller, Lincoln Mullen, and Jenny Bryan generously provided comments to improve the manuscript.
